Spending too much time at the Galactic bar: Chaotic fanning of the Ophiuchus stream

### 2016-01-14 ms.tex
comments from kathryn and nw figure

\title{Spending too much time at the Galactic bar: Chaotic stream-fanning and the Ophiuchus stream}
\author{Adrian M. Price-Whelan\altaffilmark{\colum,\adrn},
            Branimir Sesar\altaffilmark{\mpia}, 
            Kathryn V. Johnston\altaffilmark{\colum},
            Hans-Walter Rix\altaffilmark{\mpia}
}

@@

We make one modification to this method based on the idea that the Ophiuchus stream progenitor has been fully-disrupted. We add an additional parameter to the stream generation routine to specify the time of disruption, $\tau_d$. At this time, we set the offset of the Lagrange point to 0: In terms of the parametrization in \citep{fardal14}, we set $k_r = 0$ and $k_{vt}=0$ but preserve the dispersion in the release radius and velocity. Any star released after $\tau_d$ is released with a small dispersion around the progenitor orbit but no offset. This mass-loss history is intended to mimic the expected gradual evaporation of a globular cluster over a tidally-limited boundary (i.e. driven by two-body relaxation and gravitational shocks over many Gigayears) with final disruption likely to occur once the tidal boundary is less than the core radius of the cluster. The physics of this disruption is not followed exactly but rather the disruption rate and final disruption time are set by \todo{the hypothesis that the most recent shock ...}

@@

Figures~\ref{fig:mockstream0}--\ref{fig:mockstream1} show the final particle positions and line-of-sight velocities in heliocentric coordinates for the maximum likelihood mock streams (grey points) in each potential. Vertical, dashed lines show the approximate extent of the part of the stream visible in main-sequence stars (excluding the BHB stars from S15b). There are a few interesting features to note from these panels: 
\begin{enumerate}
	\item Even in the static potential (Figure~\ref{fig:mockstream0}, leftmost column), there is a slight decrease in the density for the model stream towards higher Galactic longitudes, $l$. This is a projection effect: The portion of the stream at larger $l$ is closer points almost directly towards the Sun so that the debris covers a larger area. This is more clearly seen in Figure~\ref{fig:mockstreamxyz}, which shows $x$-$z$ projections of the star particles in Galactocentric Cartesian coordinates (grey) along with the position of the Sun (symbol at $(x,z)=(-8.3,0)~{\rm kpc}$) and the ``window'' of the Heliocentric, sky-position plots of Figures~\ref{fig:mockstream0}--\ref{fig:mockstream1} (shown as blue lines).
	\item The density of the mock stream in the static potential decreases slowly rather than abruptly as is observed. This is more clearly seen in terms of the \todo{linear density along the stream, shown in Figure~\ref{fig:lineardensity}}. .... \todo{Waiting on Branimir!}
	\item The four high-dispersion BHB stars beyond the end of the stream (from S15b) don't match in position and velocity with the particle distribution from even the maximum likelihood stream model in the static potential. However, in some barred potentials the chaotic evolution of the stream stars can lead to over-densities of stars with an increased positional dispersion and significantly discrepant velocities (e.g., bar8).
	\item None of the stream models produce an appreciable density of stars with line-of-sight velocities near the S15b BHB star with the largest velocity ($\approx 320~\kms$). This star is either an interesting Ophiuchus member star or is associated with some other kinematic substructure.
	\item For the barred potentials, the stream morphology is very sensitive to the properties of the bar (especially the pattern speed) and to the initial conditions of the orbit. We have found that the morphology can vary significantly between nearby orbits in the same potential model (because these are strongly chaotic orbits), but the overall characteristics remain similar: along more strongly chaotic orbits, the debris ``fans'' more and the apparent dense part of the model streams is shorter (see, for example, Figure~\ref{fig:mockstreamxyz}, panels bar3 vs. bar8).
\end{enumerate}

@@

@@

It is still too early to say for sure that chaotic stream-fanning is occurring for the Ophiuchus stream; deeper follow-up imaging and spectroscopy over a larger area region around the stream will be needed to test the predictions of this work. With more information about the density distribution of stars in the stream and better proper motion measurements we would be able to (1) help distinguish models for the Milky Way bar independent from current methods and (2) begin to model the survivability of globular clusters orbiting in the central Milky Way \citep[e.g.,][]{gnedin97}.

@@

\begin{figure*}[p]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/mockstream0}
\caption{ Sky position, distance, and line-of-sight velocity in Heliocentric, Galactic coordinates for star particles (contours or grey points) from the maximum-likelihood mock streams in each potential. Top panels show surface density of mock stream star particles in each potential. Contours are spaced logarithmically from $10^{-2}$ to $10$ particles per sq. deg---that is, each color represents a factor of 10 difference in surface density. In the static potential (top left), the density remains high along the center of the stream, but for some of the barred potentials the density drops sharply because of chaotic stream-fanning. Vertical, dashed lines show the approximate extent of the densest part of the stream visible in main-sequence stars \citep[the segment originally detected in ][]{bernard14}.}
\label{fig:mockstream0}
\end{center}
\end{figure*}

\begin{figure*}[p]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/mockstream1}
\caption{ Same as Figure~\ref{fig:mockstream0} for the other five barred potentials. }
\label{fig:mockstream1}
\end{center}
\end{figure*}

%\begin{figure*}[p]
%\begin{center}
%\includegraphics[width=\textwidth]{figures/surface-density}
%\caption{ Surface density of mock stream star particles in each potential. Contours are spaced logarithmically from $10^{-2}$ to $10$ particles per sq. deg---that is, each color represents a factor of 10 difference in surface density. In the static potential (top left), the density remains high along the center of the stream, but for some of the barred potentials the density drops sharply because of chaotic stream-fanning.}
%\label{fig:surfacedensity}
%\end{center}
%\end{figure*}

>>>>

\title{Spending too much time at the Galactic bar: Chaotic fanning of the Ophiuchus stream}
\author{Adrian M. Price-Whelan\altaffilmark{\colum,\adrn},
            Branimir Sesar\altaffilmark{\mpia}, 
            Kathryn V. Johnston\altaffilmark{\colum},
            Hans-Walter Rix\altaffilmark{\mpia}
}

@@

We make one modification to this method based on the idea that the Ophiuchus stream progenitor has been fully-disrupted. We add an additional parameter to the stream generation routine to specify the time of disruption, $\tau_d$. At this time, we set the offset of the Lagrange point to 0: In terms of the parametrization in \citep{fardal14}, we set $k_r = 0$ and $k_{vt}=0$ but preserve the dispersion in the release radius and velocity. Any star released after $\tau_d$ is released with a small dispersion around the progenitor orbit but no offset. This mass-loss history is intended to mimic the expected gradual evaporation of a globular cluster over a tidally-limited boundary (i.e. driven by two-body relaxation and gravitational shocks over many Gigayears) with final disruption likely to occur once the tidal boundary is less than the core radius of the cluster. The physics of this disruption is not followed exactly but rather the disruption rate and final disruption time are set by the hypothesis that the most recent combined pericenter and disk shock fully disrupted the cluster, but, critically, that the cluster has been losing debris over its entire orbital history.

@@

Figures~\ref{fig:mockstream0}--\ref{fig:mockstream1} show the final particle positions and line-of-sight velocities in heliocentric coordinates for the maximum likelihood mock streams (grey points) in each potential. Vertical, dashed lines show the approximate extent of the part of the stream visible in main-sequence stars (excluding the BHB stars from S15b). There are a few interesting features to note from these panels: 
\begin{enumerate}
	\item Even in the static potential (Figure~\ref{fig:mockstream0}, leftmost column), there is a slight decrease in the density for the model stream towards higher Galactic longitudes, $l$. This is a projection effect: The portion of the stream at larger $l$ is closer points almost directly towards the Sun so that the debris covers a larger area. This is more clearly seen in Figure~\ref{fig:mockstreamxyz}, which shows $x$-$z$ projections of the star particles in Galactocentric Cartesian coordinates (grey) along with the position of the Sun (symbol at $(x,z)=(-8.3,0)~{\rm kpc}$) and the ``window'' of the Heliocentric, sky-position plots of Figures~\ref{fig:mockstream0}--\ref{fig:mockstream1} (shown as blue lines).
	\item The density of the mock stream in the static potential decreases slowly rather than abruptly as is observed. This is more clearly seen in terms of the density contrast between stream stars and background stars, visualized in Figure~\ref{fig:densitymaps}: This figure shows mock sky-density maps of stars generated by superimposing the maximum likelihood model streams over a noisy background of stars. The number of mock stream star particles used to generate the map has been normalized such that the total number of stars within the observed extent of the stream (between $5.85 > l > 3.81~{\rm deg}$) is equal to the number of stars attributed to the Ophiuchus stream in the PS1 data \citep[$N \approx 500$][]{bernard14}.
	\item The four high-dispersion BHB stars beyond the end of the stream (from S15b) don't match in position and velocity with the particle distribution from even the maximum likelihood stream model in the static potential. However, in some barred potentials the chaotic evolution of the stream stars can lead to over-densities of stars with an increased positional dispersion and significantly discrepant velocities (e.g., bar8).
	\item None of the stream models produce an appreciable density of stars with line-of-sight velocities near the S15b BHB star with the largest velocity ($\approx 320~\kms$). This star is either an interesting Ophiuchus member star or is associated with some other kinematic substructure.
	\item For the barred potentials, the stream morphology is very sensitive to the properties of the bar (especially the pattern speed) and to the initial conditions of the orbit. We have found that the morphology can vary significantly between nearby orbits in the same potential model (because these are strongly chaotic orbits), but the overall characteristics remain similar: along more strongly chaotic orbits, the debris ``fans'' more and the apparent dense part of the model streams is shorter (see, for example, Figure~\ref{fig:mockstreamxyz}, panels bar3 vs. bar8).
\end{enumerate}

@@

\todo{Add sentences about Hattori ophiuchus paper}

@@

It is still too early to say for sure that chaotic stream-fanning is occurring for the Ophiuchus stream; deeper follow-up imaging and spectroscopy over a larger area region around the stream will be needed to test the predictions of this work and compare with other possible scenarios. For example, recent work has shown that if the Ophichus stream progenitor orbit is in resonance with the bar, the debris can remain short for at least 1 Gyr \citep{hattori15}. In this model, there would be no nearby, high-dispersion debris, and the pattern speed of the bar would likely be related to the orbital frequencies of the progenitor orbit.

 With more information about the density distribution of stars in the stream and better proper motion measurements we would be able to (1) help distinguish models for the Milky Way bar independent from current methods and (2) begin to model the survivability of globular clusters orbiting in the central Milky Way \citep[e.g.,][]{gnedin97}.

@@

\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/mockstream0}
\caption{ Sky position, distance, and line-of-sight velocity in Heliocentric, Galactic coordinates for star particles (contours or grey points) from the maximum-likelihood mock streams in each potential. Top panels show surface density of mock stream star particles in each potential. Contours are spaced logarithmically from $10^{-2}$ to $10$ particles per sq. deg---that is, each color represents a factor of 10 difference in surface density. In the static potential (top left), the density remains high along the center of the stream, but for some of the barred potentials the density drops sharply because of chaotic stream-fanning. Vertical, dashed lines show the approximate extent of the densest part of the stream visible in main-sequence stars \citep[the segment originally detected in ][]{bernard14}.}
\label{fig:mockstream0}
\end{center}
\end{figure*}

\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/mockstream1}
\caption{ Same as Figure~\ref{fig:mockstream0} for the other five barred potentials. }
\label{fig:mockstream1}
\end{center}
\end{figure*}

\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/densitymaps}
\caption{ Simulated maps of the 2D density of star particles from the maximum-likelihood mock streams in each potential with a noisy background of stars binned into 10' by 10' pixels. The background star density is assumed to be Poisson with $\lambda = 42$ \citep[see Figure 3 in][where the typical background density is $\approx\frac{60}{(0.2~{\deg})^2}$]{bernard14}. The mock stream particles are down-sampled so that the total number of particles in the region of sky that the stream is seen as an over-density matches the observed number of stars \citep[$N\approx500$][]{bernard14}. Color-scale is stretched so that white to black is 2nd to 98th percentile.}
\label{fig:densitymaps}
\end{center}
\end{figure*}
